---
title: "Analyse Activity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


## Load Required Libraries

```{r  results='hide', message=FALSE, warning=FALSE}
require(dplyr)
require(ggplot2)
require(lubridate)
```

## Read the required data set for analysis and examine summary

```{r}
activity<-read.csv("activity.csv")
activity$date<-as.Date(activity$date)
summary(activity)
```

## Calculate mean total number of steps taken per day
```{r}
tot_steps_per_day<-activity %>% group_by(date) %>% summarize(total_steps=sum(steps))
```

Plot a histogram of the total number of steps taken each day and calculate mean and median of steps taken per day

```{r}
step_plot<-ggplot(tot_steps_per_day,aes(total_steps)) + 
        geom_histogram(colour = "darkgreen",fill="red") + 
        ggtitle("                            Total Steps Taken Per Day") +
        xlab("Step Count")+
        ylab("Frequency")
step_plot

mean(tot_steps_per_day$total_steps,na.rm=TRUE)
median(tot_steps_per_day$total_steps,na.rm=TRUE)

```

## Calculate average daily activity pattern

```{r}
average_5min<-activity %>% group_by(interval) %>% summarize(mean_interval_steps=mean(steps,na.rm=TRUE))
plot_interval_steps<-ggplot(average_5min,aes(interval,mean_interval_steps)) + geom_line( colour="red",size=1) +
                        ggtitle("Average Steps taken in 5 Min Interval") + xlab("Interval") +ylab("Average Steps")
plot_interval_steps

```

On average across all the days in the dataset, the following 5 minute interval contains the maximum number of steps

```{r}
max_steps<-average_5min[average_5min$mean_interval_steps==max(average_5min$mean_interval_steps),]
max_steps$interval

```


##Impute Missing Values

```{r}
#Calculate Total number of missing values
missing_values<-nrow(activity[is.na(activity$steps),])
missing_values
st_indices<-which(is.na(activity$steps))
activity2<-activity

#Impute Missing Data using mean for that 5-minute interval
activity2[st_indices,]$steps<-sapply(activity2[st_indices,]$interval,
                                    function(x){
                                                ceiling(average_5min[average_5min$interval==x,]$mean_interval_steps)
                                            }
                                    )
#Calculate the total amount of steps taken per day after imputing
tot_steps_per_day2<-activity2 %>% group_by(date) %>% summarize(total_steps=sum(steps))

```

Plot histogram of the total number of steps taken each day after imputing

```{r}
 #Overlay 2 Historgrams (one with NAs and the other with filled in Data)
hist(tot_steps_per_day2$total_steps, breaks = 30, col = "red", xlab = "Steps", main = "Total Steps Per Day")
hist(tot_steps_per_day$total_steps, breaks = 30, col = "green", xlab = "Steps", main = "Total Steps Per Day", add=T) 
legend("topleft", c("Imputed Values ", "NA"), fill = c("red","green"))
```
After calculating mean and median on the imputed data, it is observed that mean almost remains the same, but median was increased.

```{r}
#Calculate Mean and Median of steps taken per day
mean(tot_steps_per_day2$total_steps,na.rm=TRUE)
median(tot_steps_per_day2$total_steps,na.rm=TRUE)
```

## Observe differences in activity patterns between weekdays and weekends

```{r}
activity2$daytype=ifelse(wday(activity2$date) %in% c(2:6),"WEEKDAY","WEEKEND")
activity2$daytype<-as.factor(activity2$daytype)

# Calculate 5-minute intervals and the average number of steps taken, averaged across all weekday days or weekend days.
average_5min_day<-activity2 %>% group_by(daytype,interval) %>% summarize(mean_interval_steps=mean(steps,na.rm=TRUE))
plot_day<-ggplot(average_5min_day,aes(interval,mean_interval_steps)) + 
        facet_grid(daytype~.) +
        geom_line(colour="red",size=1) +
        ggtitle("Average Steps taken in 5 Min Interval") + xlab("Interval") +ylab("Average Steps")
plot_day
```
